### 2013-07-23 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
document formatting for appearance

The spectral radius of the iteration matrix is obtained by seeking its
largest eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{multline}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}\Big[4 \phi_{i-1,j} +
    4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1}
    +\\ \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}\Big]\:.
  \label{eq:iteration_stencil}
\end{multline}
Inserting the eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. To find the maximum eigenvalue,
Eq~(\ref{eq:preconditioned_eigenvalues}) is plotted as a function of
$p$ with $p=q$ in Figure~\ref{fig:diffusion_spectrum} for various
values of $\Sigma_a$.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{chapters/parallel_mc/diffusion_spectrum.png}
  \end{center}
  \caption{\textbf{Eigenvalue spectra for the diffusion equation.}}
  \label{fig:diffusion_spectrum}
\end{figure}
We find that the maximum eigenvalue exists when $p=q=0$, giving the
following for the spectral radius of the Jacobi preconditioned
iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

@@

\subsection{Numerical Experiments }
\label{subsec:numerical_experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=5in,clip]{chapters/parallel_mc/measured_spec_rad.png}
    \end{center}
    \caption{\textbf{Measured and analytic preconditioned diffusion
        operator spectral radius as a function of the absorption cross
        section to scattering cross section ratio.} \textit{Values of
        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
        computed numerically by an eigensolver while the black dashed
        data was generated by Eq~(\ref{eq:iteration_radius}).}}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

@@

To compute the leakage fraction numerically, \sn{3}{5} histories were
sampled from a uniform source of strength unity over the global
domain. At the start of a stage of histories, the number of histories
starting in the center domain was computed and as the stage
progressed, the number of histories that exited that domain was
tallied with the ratio of the two numbers providing a numerical
measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
gives the domain leakage measurements for the domain in the center of
the global grid as well as the analytic result computed by
Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
matrix spectral radius.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/leakage_variation_2.pdf}
    \end{center}
    \caption{\textbf{Measured and analytic domain leakage as a
        function of the iteration matrix spectral radius.} \textit{To
        test the behavior with respect to domain size, $n_i=50$,
        $n_i=100$,and $n_i=200$ were used. The red data was computed
        numerically by a domain-decomposed adjoint Neumann-Ulam
        implementation, the black dashed data was generated by
        Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
        approximation, and the dashed-dotted black data was generated
        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
        rational approximation.}}
    \label{fig:measured_leakage}
  \end{spacing}
\end{figure}
Again, we note good qualitative agreement between the measured and
analytic quantities but we begin to see the limits of the leakage
approximations.

To compare the quality of the two approximations, the absolute
difference between the computed leakage fraction and that generated by
the Wigner rational and mean chord approximations is plotted in
Figure~\ref{fig:leakage_error} for all domain sizes tested.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=6.0in,clip]{chapters/parallel_mc/leakage_error_2.pdf}
    \end{center}
    \caption{\textbf{Measured and analytic domain leakage absolute
        difference as a function of the iteration matrix spectral radius.}
      \textit{To test the behavior with respect to domain size,
        $n_i=50$ (green), $n_i=100$ (blue), and $n_i=200$ (red) were
        used. The dashed lines represent the difference using the Wigner
        rational approximation while the solid lines represent the
        difference using the mean-chord approximation.}}
    \label{fig:leakage_error}
  \end{spacing}
\end{figure}
From these difference results, the mean chord approximation is shown to
have a lower difference for ill-conditioned systems as compared to the
Wigner approximation while the Wigner approximation produces less
difference for more well-conditioned systems. We also note that for the
optically thick domains, the difference is likely corresponded to that
observed in Figure~\ref{fig:measured_length} for the $k$ parameter
while the large relative difference in $k$ for optically thin domains does
not affect the approximation significantly. In general, the mean chord
approximation is a better choice to estimate the leakage fraction in a
domain from the adjoint Neumann-Ulam method and except for a single
data point with $n_i=50$, the mean chord approximation yielded leakage
fractions within 0.05 of the measured results. As the domain becomes
more optically thick (with both increasing $n_i$ and decreasing
$\rho(\ve{H})$), the approximations are more accurate.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
  \end{center}
  \caption{\textbf{Multiple set strong scaling implementation
      efficiency.} \textit{Computed relative to the 16-core 1-set base
      case. Considering iterations required to converge, the
      performance of the split history cases at much better than the
      cases with history replication.}}
  \label{fig:titan_strong_ms_impeff}
\end{figure}

@@

In this chapter we have developed a domain-decomposed parallel
algorithm for Monte Carlo synthetic acceleration and explored the
qualities of the algorithm through parallel scaling studies. The
following are the significant observations and findings.

>>>>

The spectral radius of the iteration matrix is obtained by seeking its
largest eigenvalue. As with the diffusion operator, we can use the
same analysis techniques to find the eigenvalues for the iteration
matrix. We use a few simplifications by noting that if the Jacobi
preconditioned iteration matrix is $\ve{H} = \ve{I} -
\ve{M}^{-1}\ve{D}$, the we except all terms on the diagonal of the
iteration matrix to be zero such that we have the following stencil:
\begin{multline}
  \ve{H}\boldsymbol{\phi} = \frac{\alpha D}{6h^2}\Big[4 \phi_{i-1,j} +
    4 \phi_{i+1,j} + 4 \phi_{i,j-1} + 4 \phi_{i,j+1}
    +\\ \phi_{i-1,j-1} + \phi_{i-1,j+1} + \phi_{i+1,j-1} +
    \phi_{i+1,j+1}\Big]\:.
  \label{eq:iteration_stencil}
\end{multline}
Inserting the eigenfunctions defined by
Eq~(\ref{eq:eigenfunction_form}) we get:
\begin{multline}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}\Big[4 e^{-2 \pi \imath p
      h} + 4 e^{2 \pi \imath p h} + 4 e^{-2 \pi \imath q h} + 4 e^{2
      \pi \imath q h} + e^{-2 \pi \imath p h} e^{-2 \pi \imath q h}
    \\ + e^{-2 \pi \imath p h} e^{2 \pi \imath q q} + e^{2 \pi \imath
      p h} e^{-2 \pi \imath q h} + e^{2 \pi \imath p h} e^{2 \pi
      \imath q h}\Big]\:,
  \label{eq:iteration_deriv}
\end{multline}
which simplifies to:
\begin{equation}
  \lambda_{p,q}(\ve{H}) = \frac{\alpha D}{6h^2}[ 8 \cos(\pi p h) + 8
    \cos(\pi q h) + 4 \cos(\pi p h) \cos(\pi q h)]\:,
  \label{eq:iteration_spectrum}
\end{equation}
giving the eigenvalue spectrum for the Jacobi preconditioned iteration
matrix. We find that the maximum eigenvalue exists when $p=q=0$,
giving the following for the spectral radius of the Jacobi
preconditioned iteration matrix:
\begin{equation}
  \rho(\ve{H}) = \frac{10 \alpha D}{3 h^2}\:.
  \label{eq:iteration_radius}
\end{equation}

@@

\subsection{Numerical Experiments }
\label{subsec:numerical_experiments}
To test the relationships developed by the spectral analysis, we form
two simple numerical experiments using the diffusion model problem:
one to measure the length of the random walks as a function of the
iteration matrix eigenvalues, and one to measure the domain leakage
fraction as a function of the iteration matrix eigenvalues and the
discretization properties. Before doing this, we verify our
computation of the spectral radius of the iteration matrix by
numerically computing the largest eigenvalue of the diffusion operator
using an iterative eigenvalue solver. For this verification, a $100
\times 100$ square grid with $h=0.01$, $h=0.1$, and $h=1.0$ and the
absorption cross varied from 0 to 100 while the scattering cross
section was fixed at unity. Figure~\ref{fig:measured_spec_rad} gives
the measured spectral radius of the iteration matrix and the computed
spectral radius for the preconditioned diffusion operator using
Eq~(\ref{eq:iteration_radius}) as function of the absorption to
scattering ratio $(\Sigma_a / \Sigma_s)$. Excellent agreement was
observed between the analytic and numerical results with all data
points computed within the tolerance of the iterative eigenvalue
solver.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4in,clip]{chapters/parallel_mc/measured_spec_rad.png}
    \end{center}
    \caption{\textbf{Measured and analytic preconditioned diffusion
        operator spectral radius as a function of the absorption cross
        section to scattering cross section ratio.} \textit{Values of
        $h=0.01$, $h=0.1$, and $h=1.0$ were used. The red data was
        computed numerically by an eigensolver while the black dashed
        data was generated by Eq~(\ref{eq:iteration_radius}).}}
    \label{fig:measured_spec_rad}
  \end{spacing}
\end{figure}

@@

To compute the leakage fraction numerically, \sn{3}{5} histories were
sampled from a uniform source of strength unity over the global
domain. At the start of a stage of histories, the number of histories
starting in the center domain was computed and as the stage
progressed, the number of histories that exited that domain was
tallied with the ratio of the two numbers providing a numerical
measure for the leakage fraction. Figure~\ref{fig:measured_leakage}
gives the domain leakage measurements for the domain in the center of
the global grid as well as the analytic result computed by
Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
matrix spectral radius.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_variation_2.pdf}
    \end{center}
    \caption{\textbf{Measured and analytic domain leakage as a
        function of the iteration matrix spectral radius.} \textit{To
        test the behavior with respect to domain size, $n_i=50$,
        $n_i=100$,and $n_i=200$ were used. The red data was computed
        numerically by a domain-decomposed adjoint Neumann-Ulam
        implementation, the black dashed data was generated by
        Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
        approximation, and the dashed-dotted black data was generated
        by Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner
        rational approximation.}}
    \label{fig:measured_leakage}
  \end{spacing}
\end{figure}
Again, we note good qualitative agreement between the measured and
analytic quantities but we begin to see the limits of the leakage
approximations.

To compare the quality of the two approximations, the absolute
difference between the computed leakage fraction and that generated by
the Wigner rational and mean chord approximations is plotted in
Figure~\ref{fig:leakage_error} for all domain sizes tested.
\begin{figure}[t!]
  \begin{spacing}{1.0}
    \begin{center}
      \includegraphics[width=4.75in,clip]{chapters/parallel_mc/leakage_error_2.pdf}
    \end{center}
    \caption{\textbf{Measured and analytic domain leakage absolute
        difference as a function of the iteration matrix spectral radius.}
      \textit{To test the behavior with respect to domain size,
        $n_i=50$ (green), $n_i=100$ (blue), and $n_i=200$ (red) were
        used. The dashed lines represent the difference using the Wigner
        rational approximation while the solid lines represent the
        difference using the mean-chord approximation.}}
    \label{fig:leakage_error}
  \end{spacing}
\end{figure}
From these difference results, the mean chord approximation is shown to
have a lower difference for ill-conditioned systems as compared to the
Wigner approximation while the Wigner approximation produces less
difference for more well-conditioned systems. We also note that for the
optically thick domains, the difference is likely corresponded to that
observed in Figure~\ref{fig:measured_length} for the $k$ parameter
while the large relative difference in $k$ for optically thin domains does
not affect the approximation significantly. In general, the mean chord
approximation is a better choice to estimate the leakage fraction in a
domain from the adjoint Neumann-Ulam method and except for a single
data point with $n_i=50$, the mean chord approximation yielded leakage
fractions within 0.05 of the measured results. As the domain becomes
more optically thick (with both increasing $n_i$ and decreasing
$\rho(\ve{H})$), the approximations are more accurate.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/parallel_mc/titan_strong_ms_impeff.pdf}
  \end{center}
  \caption{\textbf{Multiple set strong scaling implementation
      efficiency.} \textit{Computed relative to the 16-core 1-set base
      case. Considering iterations required to converge, the
      performance of the split history cases at much better than the
      cases with history replication.}}
  \label{fig:titan_strong_ms_impeff}
\end{figure}

@@

In this chapter we have developed a domain-decomposed parallel
algorithm for Monte Carlo Synthetic Acceleration and explored the
qualities of the algorithm through parallel scaling studies. The
following are the significant observations and findings.
